Learning an Intervention via Causal Inference

### 2016-02-01 causal-bandit/icml-2015/setup.tex
Start restructure

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
The \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$. 
An edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$. 
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
An intervention ``mutilates'' the original graph by removing edges into the variables in $\vec{X}$ and the resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Further details can be found in Chapter 21 of~\cite{Koller2009}.

@@

We will consider and analyse two different goals for the learner: 1) identifying the intervention with highest expected reward by round $T$, and 2) maximising the total expected reward over all $T$ rounds. 
We denote the expected reward for the intervention $do(\vec{X} = \vec{x})$ by $\mu_{\vec{x}} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{\vec{x}} \mu_{\vec{x}}$.
Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention chosen by the learner on round $T$:
\eqn{
\label{eq:regret-simple}
\simpleregret = \mu^* - \E{\mu_{\vec{x}_T}}\,.
}
Performance for the second goal is measured by \emph{cumulative regret}, the difference between the total reward obtained if the optimal intervention were known in advance and the expected reward of the learner:
\eqn{
\label{eq:regret}
R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
}
Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
Cumulative regret is more appropriate for online agents that may continue to adapt their policy. 
We analyse both cases in \S\ref{sec:simple-regret} and \S\ref{sec:cum-regret}, respectively.

\todom{Mention pruning of actions?}
\todot{I don't understand this at all. Later we are assuming $X_1,\ldots,X_N$ are binary and there are $N$ of them}
We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.

>>>>

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from the parents of $X_i$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

@@

The learning goal we will consider and analyse is identifying the intervention with highest expected reward by round $T$. 
This objective is most appropriate when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
\todom{Citation}
We denote the expected reward for the intervention $do(\vec{X} = \vec{x})$ by $\mu_{\vec{x}} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{\vec{x}} \mu_{\vec{x}}$.
Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention chosen by the learner on round $T$:
\eqn{
\label{eq:regret-simple}
\simpleregret = \mu^* - \E{\mu_{\vec{x}_T}}\,.
}

Another common objective in bandit problems is maximising the \emph{cumulative regret}, the total expected reward over all $T$ rounds. 
We have established upper bounds for cumulative regret for parallel bandits but not the general causal bandit case.
Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.
